Bayesian statistics and swine flu

### 2009-05-11 docs/pandemic/bayesPandemic.rnw
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/diseasemapping@293 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\documentclass{article}
\usepackage{Sweave}
\title{Priors for swine flu}

\begin{document}

\maketitle

>>>>

\documentclass{beamer}
\usepackage[nogin]{Sweave}
\title{Bayesian statistics and  swine flu}
\author{Patrick Brown, Dionne Gesink Law, Peter Neal} 
\usepackage{amsmath,subfigure}

\SweaveOpts{echo=FALSE,fig=true,prefix.string=Figures/G,height=2,width=3}

\begin{document}
\begin{frame}
\maketitle

\end{frame}

<<setup,fig=false>>=
options(SweaveHooks=list(fig=function() par(mar=c(2.5,2.5,0.1,0.1), mgp=c(1.5, 0.5, 0), cex=0.6)))
# source in all the R files
Rfiles=system("CMD /K dir /B ..\\..\\pkg\\pandemic\\R\\*.R", intern=T)
Rfiles = grep("\\.R$", Rfiles, value=T)
for(D in Rfiles) {
	source(paste("..\\..\\pkg\\pandemic\\R\\", D, sep=""))
#$
}
@

\begin{frame}
\frametitle{The Problem}
\begin{itemize}
	\item Some people will get infected with swine flu
	\item Some will get hospitalized, some will die
	\item How do we fatality rate and the hospitalization rate?
	\item Problem: censoring, some individuals in the dataset have not yet had their hospitalization and/or fatality
	\item Crude rates underestimate the rates, and result in estimates increasing over time
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{The Model}

\begin{block}{Mild}
\begin{center}
 Infection $\stackrel{f_I(t)}{\Longrightarrow}$ Onset $\stackrel{f_{OM}(t;)}{\Longrightarrow}$ \\ Medical  $\stackrel{f_{MR}(t)}{\Longrightarrow}$ Recovery
\end{center}

\end{block}

\begin{block}{Serious} 
\begin{center}
Infection$\stackrel{f_I(t)}{\Longrightarrow}$ Onset $\stackrel{f_{OS}(t)}{\Longrightarrow}$ Medical $\stackrel{f_{SH}(t)}{\Longrightarrow}$\\ Hospitalization $\stackrel{f_{SR}(t;)}{\Longrightarrow}$ Recovery
\end{center}\end{block}
\begin{block}{Deadly} \begin{center}Infection $\stackrel{f_I(t)}{\Longrightarrow}$ Onset $\stackrel{f_{OD}(t)}{\Longrightarrow}$ Medical $\stackrel{f_{HD}(t)}{\Longrightarrow}$\\ Hospitalization $\stackrel{f_{D}(t)}{\Longrightarrow}$ Death 
\end{center}
\end{block}

\end{frame}

\begin{frame}
\frametitle{Distributions}
\begin{itemize}
	\item $f(t;\mu,\nu,\delta)$ is the distribution for time to an event.  
	\item \ldots a rounded, zero-inflated Weibull
\begin{align*}
Z\sim&\text{Bern}(\delta)\\
[Y|Z=1] =& 0 \text{w.p.} 1\\
pr(T=t|Z=0) =&  \text{pr}(\max(t-0.5,0) < y < t+0.5)\\
y \sim & \text{Weibull}(\mu/\Gamma(1 + 1/\nu), \nu)
\end{align*}
\item $\delta$ is the probability of transitioning immediately for an administrative (not biological) reason, 
\begin{itemize}
	\item i.e.\ arriving dead at the hospital.
\end{itemize}
\item Barring administrative zeros, $\mu = E(T|Z=0)$ is the average time to an event
\begin{itemize}
	\item $\mu_I$ is average time to infection
\end{itemize}
\item $\nu$ is a Weibull shape parameter
\begin{itemize}
	\item $\nu=1$ is constant hazards
	\item $\nu > 1$ results in more regular event times.
\end{itemize}

\end{itemize}

\end{frame}

\begin{frame}
\frametitle{Priors}
\begin{itemize}
	\item What does expert opinion tell us these parameters should be?
	\item Based on previous experience, not the current data.
	\item We need a distribution, not a single value
\begin{itemize}
	\item If we `knew' the correct value, there would be no need for this model
\end{itemize}
\item Gamma or Beta priors, parametrized by mean and standard deviation
\item Average time to onset of symptoms is expected to be 3.5 days, but the average time could be a day longer or shorter.
\begin{itemize}
	\item $E(\mu_I = 3.5)$ and $sd(\mu_I = 0.5$
	\item $\mu_I \sim \text{Gamma}(3.5^2/0.5^2 , 3.5/0.5^2 )$
\end{itemize}
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Proportion parameters}

\begin{itemize}
	\item $\rho_D$ is the fatality ratio, or probability of a deadly infection
	\item $\rho_S$ is the proportion of non-fatalities which are hospitalized
	\item so the probability of being hospitalized is $\rho_D + (1-\rho_D)\rho_S$
	\item parametrize this way to ensure the probabilities of being Mild, Severe, or Deadly add to 1.
	\item beta priors on $\rho_D$ and $\rho_S$, parametrized by mean and standard deviation
	\item i.e.\ $\rho_D$ is expected to be $0.1$, with $sd(\rho_D = 0.05)$
\end{itemize}

\end{frame}

\begin{frame}[fragile]
\frametitle{What we need}
\begin{verbatim}
transition parameter mean sd distribution
InfOns	mean	1	1	gamma	
InfOns	shape	1	1	gamma	
InfOns	zeros	0.1	0.1	beta	
OnsMedM	mean	1	1	gamma	
OnsMedM	shape	1	1	gamma	
OnsMedM	zeros	0.1	0.1	beta	
OnsMedS	mean	1	1	gamma	
OnsMedS	shape	1	1	gamma	
OnsMedS	zeros	0.1	0.1	beta	
OnsMedD	mean	1	1	gamma	
OnsMedD	shape	1	1	gamma	
OnsMedD	zeros	0.1	0.1	beta	
\end{verbatim}
\end{frame}

\begin{frame}
\frametitle{Prior graphs}
<<priorGraph, fig=false,results=tex>>=
priors= pandemicPriors()
plotPrior(priors$InfOns, file="InfOns", tex=TRUE)
@
\end{frame}

\begin{frame}
\frametitle{Inference}

\end{frame}

\begin{frame}
\frametitle{Simulated data}

\end{frame}

\begin{frame}
\frametitle{To do}
\begin{itemize}
	\item model the infection process
	\item simulate hospital loads
	\item covariates
	\item spatial infection process
\end{itemize}
\end{frame}
